A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models

### 2024-07-01 manuscript.qmd
switch back to %>% pipe; update results text to new full_simulation; fix some typos

```{r}
#| tbl-cap: Overview of software packages for power analysis for (generalized) linear mixed models in R.
#| apa-note: GUI available = whether a graphical user interface is provided (usually a Shiny app that is available online or can be downloaded to run locally), GLMM support = whether power analysis can be performed for generalized linear mixed model (otherwise only linear mixed models are supported), Design limitations = whether only a limited selection of study designs can be specified (otherwise arbitrary designs within the model class are supported).  
#| echo: false
#| message: false
# label: tbl-rpackages # There is currently a bug in apaquarto that prevents referencing for multipage tables (https://github.com/wjschne/apaquarto/issues/71)

@@

flextable::flextable(packagetable) |>
  flextable::width(j = "Package name", width = 3, unit = "cm") |>
  flextable::width(j = "Summary", width = 4, unit = "cm") |>
  flextable::width(j = "GUI available", width = 1.5, unit = "cm") |>
  flextable::width(j = "GLMM support", width = 1.5, unit = "cm") |>
  flextable::width(j = "Design limitations", width = 3, unit = "cm") |>
  flextable::width(j = "Reference", width = 2, unit = "cm") |>
  flextable::fontsize(size = 10, part = "all") |>
  flextable::theme_apa() |>
  ftExtra::colformat_md()
```
:::{.NoIndent}
*Note.* GUI available = whether a graphical user interface is provided (usually a Shiny app that is available online or can be downloaded to run locally), GLMM support = whether power analysis can be performed for generalized linear mixed model (otherwise only linear mixed models are supported), Design limitations = whether only a limited selection of study designs can be specified (otherwise arbitrary designs within the model class are supported).
:::

@@

Although the details differ for each study, every tailored simulation-based sample size planning requires a series of steps and decisions. We will introduce each step in a theoretical section, followed by the practical application based on a case study. To keep the tutorial manageable we focus on a binomial GLMM with two random intercepts and discrete predictor variables. All code in this manuscript and simulation results are available on the Open Science Framework (<https://osf.io/dhwf4/>) and on our tutorial website (<https://timo-ko.github.io/glmm_simulation_tutorial/>).

@@

$$
\begin{aligned}
& P(\text{correct diagnosis} | \text{no advice, average student, average scan}) \\
& \quad - P(\text{correct diagnosis} | \text{incorrect advice, average student, average scan})
\end{aligned}
$$
For example, the first expression is concerned with the difference between the probability that a correct diagnosis is made if correct AI advice is presented and the probability that a correct diagnosis is made if no AI advice is presented. This contrast is quantified for a hypothetical *typical* expert and a *typical* scan, where *typical* is usually defined as an average score on all attributes of the expert or scan.^[Note that a different estimand would be the so-called average treatment effect (ATE). For the ATE, the probability contrast is defined for each combination of expert and scan, and then these contrasts are averaged across all experts and scans from the target population [@lundbergWhatYourEstimand2021].].
To complete our definition of the estimand, we have to define our target population that consists of persons, stimuli, and AI advice: With respect to persons, we are only interested in experts (i.e., radiologists) and non-experts (i.e., medical students). With respect to stimuli, we are only interested in the head CT scans made from subjects that do or do not suffer from intracerebral hemorrhage. Lastly, we are only interested in AI advice given by a specific AI-enabled system.

@@

```{r}
#| echo: true
simulate <- function(n_subjects = 100, n_items = 50,
  b_0 = 0.847, b_e = 1.350, b_a = -1.253, b_c = 2.603,
  b_ea = 0.790, b_ec = -1.393,
  sd_u0s = 0.5, sd_u0i = 0.5, ...){
  require(dplyr)
  require(faux)
  # simulate design
  dat <- add_random(subject = n_subjects, item = n_items) |>
    add_between("subject", expert = c(1, 0), .prob = c(0.25, 0.75)) |>
    mutate(advice_present = rbinom(n(), 1, prob = 2/3)) |>
    mutate(advice_correct = if_else(advice_present == 1L, 
                                    rbinom(n(), 1L, prob = 0.8), 0L)) |>
    # add random effects
    add_ranef("subject", u0s = sd_u0s) |>
    add_ranef("item", u0i = sd_u0i) |>
    # compute dependent variable
    mutate(linpred = b_0 + u0i + u0s +
        b_e * expert + b_a * advice_present + b_c * advice_correct +
        b_ea * expert * advice_present + b_ec * expert * advice_correct) |>
    mutate(y_prob = plogis(linpred)) |>
    mutate(y_bin = rbinom(n = n(), size = 1, prob = y_prob))
  dat
}
```

@@

```{r}
#| label: tbl-probtable
#| tbl-cap: Assumed probabilities that an average subject solves an average item in each experimental condition.
#| apa-note: Implied equations are derived based on the model equations and setting all random intercept terms to 0.
#| echo: false
probtable <- data.frame(
  col1 = 
    c("no advice, student", "no advice, expert", "incorrect advice, student", 
      "incorrect advice, expert", "correct advice, student", "correct advice, expert"),
  col2 =
    c("0.70", "0.90", "0.40", "0.85", "0.90", "0.95"),
  col3 = 
    c("$logit(0.70) = \\beta_0$",
      "$logit(0.90) = \\beta_0 + \\beta_e$", 
      "$logit(0.40) = \\beta_0 + \\beta_a$",
      "$logit(0.85) = \\beta_0 + \\beta_e + \\beta_{a} + \\beta_{ea}$",
      "$logit(0.90) = \\beta_0 + \\beta_a + \\beta_c$",
      "$logit(0.95) = \\beta_0 + \\beta_e + \\beta_a + \\beta_c + \\beta_{ea} + \\beta_{ec}$"))
knitr::kable(probtable, col.names = c("Experimental condition",
    "$P(Y=1 \\mid \\mathbf{X=x}, \\mathbf{U} = \\mathbf{0})$",
    "Implied equation"), escape = FALSE)
```

@@

```{r}
#| echo: true
library(tidyverse)
set.seed(1)
dat <- simulate(n_subjects = 3000, n_items = 3000,
  sd_u0s = 0.5, sd_u0i = 0.5)
dat |> 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) |>
  mutate(condition = fct_recode(condition,
 "student, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
 "student, incorrect advice" = "0:1:0", "expert, incorrect advice" = "1:1:0",
 "student, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) |> 
  group_by(condition) |>
  summarize(relative_frequency = sum(y_bin) / n())
```

@@

```{r}
#| label: fig-margdist1
#| fig-cap: "Marginal distributions including means, 66% and 95% confidence intervals for all experimental conditions."
#| echo: true
library(ggdist)
dat |> 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) |>
  mutate(condition = fct_recode(condition,
"student, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
"student, incorrect advice" = "0:1:0", "expert, incorrect advice" = "1:1:0",
"student, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) |> 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45",
    breaks = "Sturges") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```
@fig-margdist1 shows the model-implied marginal distributions, including the mean, 66% and 95% intervals. We can see that, indeed, the average probabilities (black dots) slightly differ from the probabilities of average subjects and items considered previously. This difference increases with the variability of the random effects.
We can use plots like @fig-margdist1 as a useful tool to determine whether the specified standard deviations of the subject and item random intercepts ($\sigma_S$ and $\sigma_I$) are reasonable by comparing the ranges and overlap between conditions to domain knowledge.

```{r}
#| label: fig-margdist2
#| echo: false
#| fig-cap: "Marginal distributions including means, 66% and 95% confidence intervals for all experimental conditions while setting the standard deviation of item random intercepts to 0.01."
set.seed(1)
dat <- simulate(n_subjects = 3000, n_items = 3000, sd_u0i = 0.01)
dat |> 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) |>
  mutate(condition = fct_recode(condition,
"student, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
"student, incorrect advice" = "0:1:0", "expert, incorrect advice" = "1:1:0",
"student, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) |> 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45",
    breaks = "Sturges") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```
For the next plot, we have set the item standard deviation to almost zero ($\sigma_I = 0.01$). This gives us a better way to see the variability between persons. As an example, @fig-margdist2 reveals a number of implicit assumptions about the comparison between experts and students: With incorrect advice, virtually all experts have a higher probability of making a correct diagnosis compared to students when considering only items with average difficulty. In contrast, there is considerable overlap in probability between experts and students with no advice and even higher overlap with correct advice. Patterns like these should be considered carefully and discussed with the domain experts. Parameter values ($\beta$ parameters, and $\sigma_S$) should be adjusted if the implications do not seem reasonable.
We could also have a closer look at the variability between items by setting the subject standard deviation to almost zero ($\sigma_S = 0.01$).
```{r}
#| include: false
set.seed(1)
dat <- simulate(n_subjects = 3000, n_items = 3000, sd_u0s = 0.01)
dat |> 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) |>
  mutate(condition = fct_recode(condition,
"student, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
"student, incorrect advice" = "0:1:0", "expert, incorrect advice" = "1:1:0",
"student, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) |> 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45",
    breaks = "Sturges") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

The final plot demonstrates that these plots are also useful for spotting standard deviations that were specified too high. For @fig-margdist3 we have set $\sigma_S = 3$ and $\sigma_I = 3$. This implies that in each experimental condition, the probabilities that a subject solves an item are overwhelmingly close to either 0 or 1 and nothing in between, which is not a plausible assumption. These high standard deviations do not account for the inherent variability and complexity of human performance. For example, the expectation that an expert with low ability and incorrect advice would solve a difficult item with a probability close to zero is not convincing.
```{r}
#| label: fig-margdist3
#| echo: false
#| fig-cap: "Marginal distributions including means, 66% and 95% confidence intervals for all experimental conditions while setting the standard deviation of subject and item random intercepts to 3."
set.seed(1)
dat <- simulate(n_subjects = 3000, n_items = 3000, sd_u0s = 3, sd_u0i = 3)
dat |> 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) |>
  mutate(condition = fct_recode(condition,
    "student, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "student, incorrect advice" = "0:1:0", "expert, incorrect advice" = "1:1:0",
    "student, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) |> 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45",
    breaks = "Sturges") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

@@

```{r}
#| echo: true
library(tidyverse)
library(marginaleffects)
library(tinytable)
grid2 <- expand_grid(advice_present = 0:1, 
  advice_correct = 0:1, expert = 0:1)
grid2
preds <- predictions(fit, newdata = grid2, 
  type = "response", re.form = NA)
print(preds, style = "tinytable") |> theme_tt(theme = "resize")
```

@@

```{r}
#| echo: true
#| size: footnotesize
contrasts <- preds |> 
  hypotheses(hypothesis = c(
    "b8 = b2",  # (correct advice, expert) - (no advice, expert)
    "b2 = b6",  # (no advice, expert) - (incorrect advice, expert) 
    "b7 = b1",  # (correct advice, student) - (no advice, student)
    "b1 = b5"), # (no advice, student) - (incorrect advice, student)
    equivalence = c(0, 0))
print(contrasts, style = "tinytable") |> theme_tt(theme = "resize")
```

@@

```{r}
#| echo: true
sim_and_analyse <- function(
  formula_chr = "y_bin ~ 1 + expert + advice_present + advice_correct + 
    expert:advice_present + expert:advice_correct + (1|subject) + (1|item)",
  contrasts = c("b8 = b2", "b2 = b6", "b7 = b1", "b1 = b5"), ...){
  require(lme4)
  require(marginaleffects)
  require(tidyr)
  # simulate data
  dat <- simulate(...)
  # fit model
  model <- glmer(as.formula(formula_chr), data = dat, family = "binomial")
  # compute contrasts
  contr_df <- expand_grid(advice_present = 0:1, advice_correct = 0:1,
    expert = 0:1)
  predictions(model, newdata = contr_df, type = "response", re.form = NA) |>
    hypotheses(hypothesis = contrasts, equivalence = c(0, 0)) |>
    data.frame()
}
```

@@

The next code chunk specifies a simulation grid with different settings for both the number of subjects (`n_subjects`) and the number of items (`n_items`), each combination being repeated `rep` times.
The more repetitions, the more accurately power and precision can be estimated. We chose 300 repetitions for the data simulation at hand to strike a balance between achieving a robust estimate and remaining computationally feasible.
With the current settings, this simulation takes about one hour on a MacBook Pro from 2020 with M1 chip and 16 GB working memory. If you want to quickly experiment with the code yourself, a setting with `workers = 4` and `rep = 5` should finish in less than 5 minutes, even on smaller machines.

```{r}
#| eval: false
#| echo: true
library(furrr)
sim_result <- crossing(
  rep = 1:300,
  n_subjects = c(100, 150, 200, 250),
  n_items = c(10, 30, 50, 70)
) |>
  mutate(res = future_pmap(., sim_and_analyse, 
    .options = furrr_options(seed = TRUE))) |>
  unnest(col = res)
```

@@

```{r}
#| label: fig-finalpwr
#| fig-cap: "Simulation-based power estimates including 95% confidence interval of the case study for different numbers of subjects and items, based on a significance level of 0.05."
#| echo: true
library(binom)
alpha <- 0.05
power <- sim_result |>
  pivot_wider(names_from = term, names_sep = "_", 
    values_from = estimate:p.value.equiv) |>
  group_by(n_subjects, n_items) |> 
  summarise(
    power = mean(`p.value.noninf_b1=b5` < alpha & 
        `p.value.noninf_b8=b2` < alpha & `p.value.noninf_b2=b6` < alpha & 
        `p.value.noninf_b7=b1` < alpha), 
    n_sig = sum(`p.value.noninf_b1=b5` < alpha & 
        `p.value.noninf_b8=b2` < alpha & `p.value.noninf_b2=b6` < alpha & 
        `p.value.noninf_b7=b1` < alpha),
    n = n(),
    ci.lwr = binom.confint(n_sig, n, method = "wilson")$lower,
    ci.upr = binom.confint(n_sig, n, method = "wilson")$upper, 
    .groups = "drop")
power |>
  mutate(across(c(n_subjects, n_items), factor)) |>
  ggplot(aes(n_subjects, n_items, fill = power)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f \n [%.2f; %.2f]", 
                                power, ci.lwr, ci.upr)), 
    color = "white", size = 4) +
  scale_fill_viridis_c(limits = c(0, 1)) +
  xlab("number of subjects") + ylab("number of items")
```

@@

When target power has been specified, the number of subjects and the number of items in our study design can be traded against each other based on practical considerations. Although it is recommended to justify the power [@lakensSampleSizeJustification2022], we adopt the common heuristic to target a power of 0.8 to detect an effect of the expected size implied by our data simulation. This could be achieved by collecting data from 200 subjects (about 25% of which will be experts), each completing the same 50 items (with advice present in about 67% of cases, which is correct in about 80% of cases with present advice). If collecting data from 200 subjects is not feasible, an alternative would be to recruit 150 subjects but increase the length of the experiment to over 70 items. However, 70 items might take too long to complete for the radiologists participating in the study, who have a busy schedule. The simulation suggests that it might also be possible to plan a shorter experiment with only 30 items if it is feasible to recruit an even higher number of subjects (\> 250, to be determined by additional rounds of power analysis). Design parameters that also affect power, and which could be investigated in the simulation to find a more optimal trade-off, are the ratio of experts to students, the frequency of whether advice is presented at all, and whether it is correct or incorrect.^[An advanced alternative to our heuristic trade-off between subjects and items would be to explicitly optimize for cost-efficient study designs using the *mlpwr* package [@zimmerSampleSizePlanning2023]. The package performs simulation based power analysis in a surrogate modeling framework that can evaluate power for a large grid of design combinations more efficiently. The framework also requires users to write their own `sim_and_analyse` function and can be considered a great extension on what we cover in our tutorial.]

>>>>

```{r}
#| tbl-cap: Table 1: Overview of software packages for power analysis for (generalized) linear mixed models in R.
#| apa-note: GUI available = whether a graphical user interface is provided (usually a Shiny app that is available online or can be downloaded to run locally), GLMM support = whether power analysis can be performed for generalized linear mixed model (otherwise only linear mixed models are supported), Design limitations = whether only a limited selection of study designs can be specified (otherwise arbitrary designs within the model class are supported).  
#| echo: false
#| message: false
# label: tbl-rpackages # There is currently a bug in apaquarto that prevents referencing for multipage tables (https://github.com/wjschne/apaquarto/issues/71)

@@

flextable::flextable(packagetable) %>%
  flextable::width(j = "Package name", width = 3, unit = "cm") %>%
  flextable::width(j = "Summary", width = 4, unit = "cm") %>%
  flextable::width(j = "GUI available", width = 1.5, unit = "cm") %>%
  flextable::width(j = "GLMM support", width = 1.5, unit = "cm") %>%
  flextable::width(j = "Design limitations", width = 3, unit = "cm") %>%
  flextable::width(j = "Reference", width = 2, unit = "cm") %>%
  flextable::fontsize(size = 10, part = "all") %>%
  flextable::theme_apa() %>%
  ftExtra::colformat_md()
```
:::{.NoIndent}
*Note.* GUI available = whether a graphical user interface is provided (usually a Shiny app that is available online or can be downloaded to run locally), GLMM support = whether power analysis can be performed for generalized linear mixed model (otherwise only linear mixed models are supported), Design limitations = whether only a limited selection of study designs can be specified (otherwise arbitrary designs within the model class are supported).
:::

@@

Although the details differ for each study, every tailored simulation-based sample size planning requires a series of steps and decisions. We will introduce each step in a theoretical section, followed by the practical application based on a case study. To keep the tutorial manageable, we focus on a binomial GLMM with two random intercepts and discrete predictor variables. All code in this manuscript and simulation results are available on the Open Science Framework (<https://osf.io/dhwf4/>) and on our tutorial website (<https://timo-ko.github.io/glmm_simulation_tutorial/>).

@@

$$
\begin{aligned}
& P(\text{correct diagnosis} | \text{no advice, average student, average scan}) \\
& \quad - P(\text{correct diagnosis} | \text{incorrect advice, average student, average scan})
\end{aligned}
$$
For example, the first expression is concerned with the difference between the probability that a correct diagnosis is made if correct AI advice is presented and the probability that a correct diagnosis is made if no AI advice is presented. This contrast is quantified for a hypothetical *typical* expert and a *typical* scan, where *typical* is usually defined as an average score on all attributes of the expert or scan.^[Note that a different estimand would be the so-called average treatment effect (ATE). For the ATE, the probability contrast is defined for each combination of expert and scan, and then these contrasts are averaged across all experts and scans from the target population [@lundbergWhatYourEstimand2021].]
To complete our definition of the estimand, we have to define our target population that consists of persons, stimuli, and AI advice: With respect to persons, we are only interested in experts (i.e., radiologists) and non-experts (i.e., medical students). With respect to stimuli, we are only interested in the head CT scans made from subjects that do or do not suffer from intracerebral hemorrhage. Lastly, we are only interested in AI advice given by a specific AI-enabled system.

@@

```{r}
#| echo: true
simulate <- function(n_subjects = 100, n_items = 50,
  b_0 = 0.847, b_e = 1.350, b_a = -1.253, b_c = 2.603,
  b_ea = 0.790, b_ec = -1.393,
  sd_u0s = 0.5, sd_u0i = 0.5, ...){
  require(dplyr)
  require(faux)
  # simulate design
  dat <- add_random(subject = n_subjects, item = n_items) %>%
    add_between("subject", expert = c(1, 0), .prob = c(0.25, 0.75)) %>%
    mutate(advice_present = rbinom(n(), 1, prob = 2/3)) %>%
    mutate(advice_correct = if_else(advice_present == 1L, 
                                    rbinom(n(), 1L, prob = 0.8), 0L)) %>%
    # add random effects
    add_ranef("subject", u0s = sd_u0s) %>%
    add_ranef("item", u0i = sd_u0i) %>%
    # compute dependent variable
    mutate(linpred = b_0 + u0i + u0s +
      b_e * expert + b_a * advice_present + b_c * advice_correct +
      b_ea * expert * advice_present + b_ec * expert * advice_correct) %>%
    mutate(y_prob = plogis(linpred)) %>%
    mutate(y_bin = rbinom(n = n(), size = 1, prob = y_prob))
  dat
}
```

@@

```{r}
#| tbl-cap: Table 2: Assumed probabilities that an average subject solves an average item in each experimental condition.
#| apa-note: Implied equations are derived based on the model equations and setting all random intercept terms to 0.
#| echo: false
probtable <- data.frame(
  col1 = 
    c("no advice, student", "no advice, expert", "incorrect advice, student", 
      "incorrect advice, expert", "correct advice, student", "correct advice, expert"),
  col2 =
    c("0.70", "0.90", "0.40", "0.85", "0.90", "0.95"),
  col3 = 
    c("$logit(0.70) = \\beta_0$",
      "$logit(0.90) = \\beta_0 + \\beta_e$", 
      "$logit(0.40) = \\beta_0 + \\beta_a$",
      "$logit(0.85) = \\beta_0 + \\beta_e + \\beta_{a} + \\beta_{ea}$",
      "$logit(0.90) = \\beta_0 + \\beta_a + \\beta_c$",
      "$logit(0.95) = \\beta_0 + \\beta_e + \\beta_a + \\beta_c + \\beta_{ea} + \\beta_{ec}$"))
knitr::kable(probtable, col.names = c("Experimental condition",
    "$P(Y=1 \\mid \\mathbf{X=x}, \\mathbf{U} = \\mathbf{0})$",
    "Implied equation"), escape = FALSE)
```

@@

```{r}
#| echo: true
library(tidyverse)
set.seed(1)
dat <- simulate(n_subjects = 3000, n_items = 3000,
  sd_u0s = 0.5, sd_u0i = 0.5)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
 "student, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
 "student, incorrect advice" = "0:1:0", "expert, incorrect advice" = "1:1:0",
 "student, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  group_by(condition) %>%
  summarize(relative_frequency = sum(y_bin) / n())
```

@@

```{r}
#| label: fig-margdist1
#| fig-cap: "Marginal distributions including means, 66% and 95% confidence intervals for all experimental conditions."
#| echo: true
library(ggdist)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
"student, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
"student, incorrect advice" = "0:1:0", "expert, incorrect advice" = "1:1:0",
"student, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45",
    breaks = "Sturges") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```
@fig-margdist1 shows the model-implied marginal distributions, including the mean, 66% and 95% intervals. We can see that, indeed, the average probabilities (black dots) slightly differ from the probabilities of average subjects and items considered previously. This difference increases with the variability of the random effects.
We can use plots like @fig-margdist1 as a useful tool to determine whether the specified standard deviations of the subject and item random intercepts ($\sigma_S$ and $\sigma_I$) are reasonable by comparing the ranges and overlap between conditions to domain knowledge.

```{r}
#| label: fig-margdist2
#| echo: false
#| fig-cap: "Marginal distributions including means, 66% and 95% confidence intervals for all experimental conditions while setting the standard deviation of item random intercepts to 0.01."
set.seed(1)
dat <- simulate(n_subjects = 3000, n_items = 3000, sd_u0i = 0.01)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
"student, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
"student, incorrect advice" = "0:1:0", "expert, incorrect advice" = "1:1:0",
"student, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45",
    breaks = "Sturges") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```
For the next plot, we have set the item standard deviation to almost zero ($\sigma_I = 0.01$). This gives us a better way to see the variability between persons. As an example, @fig-margdist2 reveals a number of implicit assumptions about the comparison between experts and students: With incorrect advice, virtually all experts have a higher probability of making a correct diagnosis compared to students when considering only items with average difficulty. In contrast, there is considerable overlap in probability between experts and students with no advice and even higher overlap with correct advice. Patterns like these should be considered carefully and discussed with the domain experts. Parameter values ($\beta$ parameters, and $\sigma_S$) should be adjusted if the implications do not seem reasonable.
We could also have a closer look at the variability between items by setting the subject standard deviation to almost zero ($\sigma_S = 0.01$).
```{r}
#| include: false
set.seed(1)
dat <- simulate(n_subjects = 3000, n_items = 3000, sd_u0s = 0.01)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
"student, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
"student, incorrect advice" = "0:1:0", "expert, incorrect advice" = "1:1:0",
"student, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45",
    breaks = "Sturges") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

The final plot demonstrates that these plots are also useful for spotting standard deviations that were specified too high. For @fig-margdist3 we have set $\sigma_S = 3$ and $\sigma_I = 3$. This implies that in each experimental condition, the probabilities that a subject solves an item are overwhelmingly close to either 0 or 1 and nothing in between, which is not a plausible assumption. These high standard deviations do not account for the inherent variability and complexity of human performance. For example, the expectation that an expert with low ability and incorrect advice would solve a difficult item with a probability close to zero is not convincing.
```{r}
#| label: fig-margdist3
#| echo: false
#| fig-cap: "Marginal distributions including means, 66% and 95% confidence intervals for all experimental conditions while setting the standard deviation of subject and item random intercepts to 3."
set.seed(1)
dat <- simulate(n_subjects = 3000, n_items = 3000, sd_u0s = 3, sd_u0i = 3)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
"student, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
"student, incorrect advice" = "0:1:0", "expert, incorrect advice" = "1:1:0",
"student, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45",
    breaks = "Sturges") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

@@

```{r}
#| echo: true
library(tidyverse)
library(marginaleffects)
library(tinytable)
grid2 <- expand_grid(advice_present = 0:1, 
  advice_correct = 0:1, expert = 0:1)
grid2
preds <- predictions(fit, newdata = grid2, 
  type = "response", re.form = NA)
print(preds, style = "tinytable") %>% theme_tt(theme = "resize")
```

@@

```{r}
#| echo: true
#| size: footnotesize
contrasts <- preds %>% 
  hypotheses(hypothesis = c(
    "b8 = b2",  # (correct advice, expert) - (no advice, expert)
    "b2 = b6",  # (no advice, expert) - (incorrect advice, expert) 
    "b7 = b1",  # (correct advice, student) - (no advice, student)
    "b1 = b5"), # (no advice, student) - (incorrect advice, student)
    equivalence = c(0, 0))
print(contrasts, style = "tinytable") %>% theme_tt(theme = "resize")
```

@@

```{r}
#| echo: true
sim_and_analyse <- function(
  formula_chr = "y_bin ~ 1 + expert + advice_present + advice_correct + 
    expert:advice_present + expert:advice_correct + (1|subject) + (1|item)",
  contrasts = c("b8 = b2", "b2 = b6", "b7 = b1", "b1 = b5"), ...){
  require(lme4)
  require(marginaleffects)
  require(tidyr)
  # simulate data
  dat <- simulate(...)
  # fit model
  model <- glmer(as.formula(formula_chr), data = dat, family = "binomial")
  # compute contrasts
  contr_df <- expand_grid(advice_present = 0:1, advice_correct = 0:1,
    expert = 0:1)
  predictions(model, newdata = contr_df, type = "response", re.form = NA) %>%
    hypotheses(hypothesis = contrasts, equivalence = c(0, 0)) %>%
    data.frame()
}
```

@@

The next code chunk specifies a simulation grid with different settings for both the number of subjects (`n_subjects`) and the number of items (`n_items`), each combination being repeated `rep` times.
The more repetitions, the more accurately power and precision can be estimated. We chose 300 repetitions for the data simulation at hand to strike a balance between achieving a robust estimate and remaining computationally feasible.
With the current settings, this simulation takes several hours on a MacBook Pro from 2020 with M1 chip and 16 GB working memory. If you want to quickly experiment with the code yourself, a setting with `workers = 4` and `rep = 5` should finish in less than 5 minutes, even on smaller machines.

```{r}
#| eval: false
#| echo: true
library(furrr)
sim_result <- crossing(
  rep = 1:500,
  n_subjects = c(100, 150, 200, 250),
  n_items = c(10, 30, 50, 70)
) %>%
  mutate(res = future_pmap(., sim_and_analyse, 
    .options = furrr_options(seed = TRUE))) %>%
  unnest(col = res)
```

@@

```{r}
#| label: fig-finalpwr
#| fig-cap: "Simulation-based power estimates including 95% confidence interval of the case study for different numbers of subjects and items, based on a significance level of 0.05."
#| echo: true
library(binom)
alpha <- 0.05
power <- sim_result %>%
  pivot_wider(names_from = term, names_sep = "_", 
    values_from = estimate:p.value.equiv) %>%
  group_by(n_subjects, n_items) %>% 
  summarise(
    power = mean(`p.value.noninf_b1=b5` < alpha & 
        `p.value.noninf_b8=b2` < alpha & `p.value.noninf_b2=b6` < alpha & 
        `p.value.noninf_b7=b1` < alpha), 
    n_sig = sum(`p.value.noninf_b1=b5` < alpha & 
        `p.value.noninf_b8=b2` < alpha & `p.value.noninf_b2=b6` < alpha & 
        `p.value.noninf_b7=b1` < alpha),
    n = n(),
    ci.lwr = binom.confint(n_sig, n, method = "wilson")$lower,
    ci.upr = binom.confint(n_sig, n, method = "wilson")$upper, 
    .groups = "drop")
power %>%
  mutate(across(c(n_subjects, n_items), factor)) %>%
  ggplot(aes(n_subjects, n_items, fill = power)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f \n [%.2f; %.2f]", 
                                power, ci.lwr, ci.upr)), 
    color = "white", size = 4) +
  scale_fill_viridis_c(limits = c(0, 1)) +
  xlab("number of subjects") + ylab("number of items")
```

@@

When target power has been specified, the number of subjects and the number of items in our study design can be traded against each other based on practical considerations. Although it is recommended to justify the power [@lakensSampleSizeJustification2022], we adopt the common heuristic to target a power of 0.8 to detect an effect of the expected size implied by our data simulation. This could be achieved by collecting data from 250 subjects (about 25% of which will be experts), each completing the same 50 items (with advice present in about 67% of cases, which is correct in about 80% of cases with present advice). If collecting data from 250 subjects is not feasible, an alternative would be to recruit 200 subjects but increase the length of the experiment to 70 items. However, 70 items might take too long to complete for the radiologists participating in the study, who have a busy schedule. The simulation suggests that it might also be possible to plan a shorter experiment with only 30 items if it is feasible to recruit an even higher number of subjects (to be determined by additional rounds of power analysis). Design parameters that also affect power, and which could be investigated in the simulation to find a more optimal trade-off, are the ratio of experts to students, the frequency of whether advice is presented at all, and whether it is correct or incorrect.^[An advanced alternative to our heuristic trade-off between subjects and items would be to explicitly optimize for cost-efficient study designs using the *mlpwr* package [@zimmerSampleSizePlanning2023]. The package performs simulation based power analysis in a surrogate modeling framework that can evaluate power for a large grid of design combinations more efficiently. The framework also requires users to write their own `sim_and_analyse` function and can be considered a great extension on what we cover in our tutorial.]
